---
title: "Capstone Data Wrangling"
author: "Will Pauls"
date: "February 18, 2019"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(Lahman)
library(dplyr)
library(tidyr)
library(readr)
```

## Data Wrangling 

For my Capstone project I will be pulling from the Teams and Salaries datasets from the Lahman package. Performing data wrangling tasks and merging them into a new dataset called baseball

## Steps
1. Load in Teams dataset with necessary variables
```{r, warning=FALSE, message=FALSE}

# make team data frame with necessary variables
teams = Teams
teams = subset(teams, select = c(yearID, lgID, teamID, Rank, G, W, L, DivWin, WCWin)) 
```
2. Since Salary info dates back only to 1985 I filtered the Teams dataset from 1985 and on to match 
```{r, warning=FALSE}
# grab teams from 1985 and on to match with years of payroll information
teams = filter(teams, yearID >=1985)
```
3. I added a column in teams for winning %
```{r}
# add winning % column
teams = mutate(teams, winpercent = round((W/(W+L)), digits = 3))
```
4. I added a column with a boolean data type (Y/N) of whether that team made the playoffs
```{r}
# add a playoffs column
teams$DivWin[is.na(teams$DivWin)] <- "N"
teams$WCWin[is.na(teams$WCWin)] <- "N"
teams <- mutate(teams, playoffs = if_else(DivWin == 'Y' | WCWin == 'Y', 'Y', 'N'))
teams = subset(teams, select = -c(DivWin, WCWin))
```
5. I loaded in the Salaries dataset
```{r, warning=FALSE, message=FALSE}
# add salaries to a data frame
sals = read_csv("Salaries.csv") 
```
6. I summed up team payroll by grouping by year and team
```{r}
# Get team payroll for each year
sals <- sals %>% 
  group_by(yearID, teamID) %>% 
    summarise(payroll = sum(salary))

```
7. I standardized team names in each data set to prepare for merge
```{r}
#Standardize team names 
sals$teamID[sals$teamID=="CHN"] <- "CHC"
sals$teamID[sals$teamID=="CHA"] <- "CHW"
sals$teamID[sals$teamID=="KCA"] <- "KCR"
sals$teamID[sals$teamID=="LAN"] <- "LAD"
sals$teamID[sals$teamID=="NYN"] <- "NYM"
sals$teamID[sals$teamID=="NYA"] <- "NYY"
sals$teamID[sals$teamID=="SDN"] <- "SDP"
sals$teamID[sals$teamID=="SFN"] <- "SFG"
sals$teamID[sals$teamID=="SLN"] <- "STL"
sals$teamID[sals$teamID=="TBA"] <- "TBR"
sals$teamID[sals$teamID=="WSN"] <- "WAS"

teams$teamID <- as.character(teams$teamID)
teams$teamID[teams$teamID=="CHN"] <- "CHC"
teams$teamID[teams$teamID=="CHN"] <- "CHC"
teams$teamID[teams$teamID=="CHA"] <- "CHW"
teams$teamID[teams$teamID=="KCA"] <- "KCR"
teams$teamID[teams$teamID=="LAN"] <- "LAD"
teams$teamID[teams$teamID=="NYN"] <- "NYM"
teams$teamID[teams$teamID=="NYA"] <- "NYY"
teams$teamID[teams$teamID=="SDN"] <- "SDP"
teams$teamID[teams$teamID=="SFN"] <- "SFG"
teams$teamID[teams$teamID=="SLN"] <- "STL"
teams$teamID[teams$teamID=="TBA"] <- "TBR"
teams$teamID[teams$teamID=="WSN"] <- "WAS"
teams$teamID <- as.factor(teams$teamID)
```
8. I merged the teams and salaries datasets into a new one I named baseball
```{r}
# merge data into baseball
baseball <- merge(teams, sals, by=c("yearID", "teamID"))

baseball$teamID[baseball$teamID=="ML4"] <- "MIL"
baseball$yearID <- as.factor(baseball$yearID)
baseball$playoffs <- as.factor(baseball$playoffs)
```
9. I created a column in baseball calculating the amount of payroll per team win (Payroll/W)
```{r}
# calculate Amount of payroll per team win
baseball <- baseball %>% mutate(dolperwin = payroll/W)
```
10. I calculated the average payroll in the MLB per year by summing each teams payroll by year and dividing by the number of teams for that year and added it to a new column in baseball
```{r}
#calculate total payroll per year for entire MLB
mlb <- baseball %>% 
  group_by(yearID) %>% 
     summarise(mlbpayroll = sum(as.numeric(payroll)))

# get number of teams for each year
 numteams <-  baseball %>% 
  group_by(yearID) %>%
  summarise(numteams =length(yearID))

mlb$numteams <- numteams$numteams

# calculate average payroll per year
mlb <- mlb %>% 
  mutate(avgpayroll = mlbpayroll/numteams)
mlb <- subset(mlb, select = -c(numteams))

# merge into baseball data set
baseball <- merge(baseball, mlb, by="yearID")
```
11. I divided each teams payroll by the the total MLB payroll to get a % of payroll each team has compared to entire league and added it into a new column
```{r}
# calculate percentange of total payroll each team has
baseball <- baseball %>% 
  mutate(percentofmlb = round((payroll/mlbpayroll), digits = 4))

```
12. I added a column that calculated the average amount of MLB payroll per each win
```{r}
#calculate average dollars of mlb payroll per win
wins <- baseball %>% 
  group_by(yearID) %>% 
  summarise(mlbwin= (sum(W)))

baseball <- merge(baseball, wins, by="yearID")
baseball <- baseball %>% 
  mutate(mlbdolperwin = mlbpayroll/mlbwin)

baseball <- baseball %>% group_by(yearID) %>% mutate(payrank = dense_rank((desc(payroll))))

```
13. I added a column with the difference in payroll from the previous year for each team
```{r}
# calculate teams difference in payroll from previous year
baseball<- baseball %>%
  arrange(teamID, yearID) %>%
  group_by(teamID) %>%
  mutate(paydiff = payroll - lag(payroll)) 
```
14. I created a % change of that difference in a new column
```{r}
# +/- % change of payroll from previous year
baseball<- baseball %>%
  arrange(teamID, yearID) %>%
  group_by(teamID) %>%
  mutate(percentdiff = paydiff/lag(payroll)) 

baseball$percentdiff <- round(baseball$percentdiff, 4)
```
15. I created a column with a difference in winning % from the previous year for each team
```{r}
## calculate change in winning percentage from previous year
baseball<- baseball %>%
  arrange(teamID, yearID) %>%
  group_by(teamID) %>%
  mutate(winpercentdiff = winpercent - lag(winpercent)) 
```
16. I created a column with a difference in a teams % of MLB payroll from previous year
```{r}
## calculate change in percent of mlb from previous year
baseball<- baseball %>%
  arrange(teamID, yearID) %>%
  group_by(teamID) %>%
  mutate(percentofmlbdiff = percentofmlb - lag(percentofmlb)) 
```
17. I created a column with the differencr in Payroll Rank from the previous year
```{r}
## calculate change in payroll rank from previous year
baseball<- baseball %>%
  arrange(teamID, yearID) %>%
  group_by(teamID) %>%
  mutate(payrankdiff = payrank - lag(payrank)) 
```
18. I created a column to test if payroll increased for each team that year
```{r, warning=FALSE}
# create column to test if payroll increased for team
baseball$payincreased[baseball$paydiff<=0]<- "No"
baseball$payincreased[baseball$paydiff>0]<- "Yes"

baseball$payincreased <- as.factor(baseball$payincreased)
```
19. I created a column to test if winning % increased for each team that year
```{r, warning=FALSE}
# create a column to test if winning % increased for team
baseball$winincreased[baseball$winpercentdiff<=0]<- "No"
baseball$winincreased[baseball$winpercentdiff>0]<- "Yes"

baseball$winincreased <- as.factor(baseball$winincreased)
```
20. I created a column to test if teams % of MLB payroll increased from previous year
```{r, warning=FALSE}
# create a column to test if percent of mlb increased for team
baseball$percentofmlbincreased[baseball$percentofmlbdiff<=0]<- "No"
baseball$percentofmlbincreased[baseball$percentofmlbdiff>0]<- "Yes"

baseball$percentofmlbincreased <- as.factor(baseball$percentofmlbincreased)
```
21. I created a column to test if a Teams Payroll rank increased from previous year
```{r, warning=FALSE}
# create column to test if payroll increased for team
baseball$payrankincreased[baseball$payrankdiff<=0]<- "No"
baseball$payrankincreased[baseball$payrankdiff>0]<- "Yes"

baseball$payrankincreased <- as.factor(baseball$payrankincreased)
```