---
title: "Probability and Statistics"
author: "Will Pauls"
date: "February 19, 2019"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(Lahman)
library(dplyr)
library(tidyr)
library(readr)
library(car)
library(MASS)
library(gvlma)
library(ggplot2)

# make team data frame with necessary variables
teams = Teams
teams = subset(teams, select = c(yearID, lgID, teamID, Rank, G, W, L, DivWin, WCWin)) 

# grab teams from 1985 and on to match with years of payroll information
teams = filter(teams, yearID >=1985)

# add winning % column
teams = mutate(teams, winpercent = round((W/(W+L)), digits = 3))

# add a playoffs column
teams$DivWin[is.na(teams$DivWin)] <- "N"
teams$WCWin[is.na(teams$WCWin)] <- "N"
teams <- mutate(teams, playoffs = if_else(DivWin == 'Y' | WCWin == 'Y', 'Y', 'N'))
teams = subset(teams, select = -c(DivWin, WCWin))
# add salaries to a data frame
sals = read_csv("Salaries.csv")          

# Get team payroll for each year
sals <- sals %>% 
  group_by(yearID, teamID) %>% 
    summarise(payroll = sum(salary))

#Standardize team names 
sals$teamID[sals$teamID=="CHN"] <- "CHC"
sals$teamID[sals$teamID=="CHA"] <- "CHW"
sals$teamID[sals$teamID=="KCA"] <- "KCR"
sals$teamID[sals$teamID=="LAN"] <- "LAD"
sals$teamID[sals$teamID=="NYN"] <- "NYM"
sals$teamID[sals$teamID=="NYA"] <- "NYY"
sals$teamID[sals$teamID=="SDN"] <- "SDP"
sals$teamID[sals$teamID=="SFN"] <- "SFG"
sals$teamID[sals$teamID=="SLN"] <- "STL"
sals$teamID[sals$teamID=="TBA"] <- "TBR"
sals$teamID[sals$teamID=="WSN"] <- "WAS"

teams$teamID <- as.character(teams$teamID)
teams$teamID[teams$teamID=="CHN"] <- "CHC"
teams$teamID[teams$teamID=="CHN"] <- "CHC"
teams$teamID[teams$teamID=="CHA"] <- "CHW"
teams$teamID[teams$teamID=="KCA"] <- "KCR"
teams$teamID[teams$teamID=="LAN"] <- "LAD"
teams$teamID[teams$teamID=="NYN"] <- "NYM"
teams$teamID[teams$teamID=="NYA"] <- "NYY"
teams$teamID[teams$teamID=="SDN"] <- "SDP"
teams$teamID[teams$teamID=="SFN"] <- "SFG"
teams$teamID[teams$teamID=="SLN"] <- "STL"
teams$teamID[teams$teamID=="TBA"] <- "TBR"
teams$teamID[teams$teamID=="WSN"] <- "WAS"
teams$teamID <- as.factor(teams$teamID)

# merge data into baseball
baseball <- merge(teams, sals, by=c("yearID", "teamID"))

baseball$teamID[baseball$teamID=="ML4"] <- "MIL"
baseball$yearID <- as.factor(baseball$yearID)
baseball$playoffs <- as.factor(baseball$playoffs)

# calculate Amount of payroll per team win
baseball <- baseball %>% mutate(dolperwin = payroll/W)

#calculate total payroll per year for entire MLB
mlb <- baseball %>% 
  group_by(yearID) %>% 
     summarise(mlbpayroll = sum(as.numeric(payroll)))

# get number of teams for each year
 numteams <-  baseball %>% 
  group_by(yearID) %>%
  summarise(numteams =length(yearID))

mlb$numteams <- numteams$numteams

# calculate average payroll per year
mlb <- mlb %>% 
  mutate(avgpayroll = mlbpayroll/numteams)
mlb <- subset(mlb, select = -c(numteams))

# merge into baseball data set
baseball <- merge(baseball, mlb, by="yearID")

# calculate percentange of total payroll each team has
baseball <- baseball %>% 
  mutate(percentofmlb = round((payroll/mlbpayroll), digits = 4))

#calculate average dollars of mlb payroll per win
wins <- baseball %>% 
  group_by(yearID) %>% 
  summarise(mlbwin= (sum(W)))

baseball <- merge(baseball, wins, by="yearID")
baseball <- baseball %>% 
  mutate(mlbdolperwin = mlbpayroll/mlbwin)

baseball <- baseball %>% group_by(yearID) %>% mutate(payrank = dense_rank((desc(payroll))))

# calculate teams difference in payroll from previous year
baseball<- baseball %>%
  arrange(teamID, yearID) %>%
  group_by(teamID) %>%
  mutate(paydiff = payroll - lag(payroll)) 

# +/- % change of payroll from previous year
baseball<- baseball %>%
  arrange(teamID, yearID) %>%
  group_by(teamID) %>%
  mutate(percentdiff = paydiff/lag(payroll)) 

baseball$percentdiff <- round(baseball$percentdiff, 4)

## calculate change in winning percentage from previous year
baseball<- baseball %>%
  arrange(teamID, yearID) %>%
  group_by(teamID) %>%
  mutate(winpercentdiff = winpercent - lag(winpercent)) 

## calculate change in percent of mlb from previous year
baseball<- baseball %>%
  arrange(teamID, yearID) %>%
  group_by(teamID) %>%
  mutate(percentofmlbdiff = percentofmlb - lag(percentofmlb)) 

## calculate change in payroll rank from previous year
baseball<- baseball %>%
  arrange(teamID, yearID) %>%
  group_by(teamID) %>%
  mutate(payrankdiff = payrank - lag(payrank)) 

# create column to test if payroll increased for team
baseball$payincreased[baseball$paydiff<=0]<- "No"
baseball$payincreased[baseball$paydiff>0]<- "Yes"

baseball$payincreased <- as.factor(baseball$payincreased)

# create a column to test if winning % increased for team
baseball$winincreased[baseball$winpercentdiff<=0]<- "No"
baseball$winincreased[baseball$winpercentdiff>0]<- "Yes"

baseball$winincreased <- as.factor(baseball$winincreased)

# create a column to test if percent of mlb increased for team
baseball$percentofmlbincreased[baseball$percentofmlbdiff<=0]<- "No"
baseball$percentofmlbincreased[baseball$percentofmlbdiff>0]<- "Yes"

baseball$percentofmlbincreased <- as.factor(baseball$percentofmlbincreased)

# create column to test if payroll increased for team
baseball$payrankincreased[baseball$payrankdiff<=0]<- "No"
baseball$payrankincreased[baseball$payrankdiff>0]<- "Yes"

baseball$payrankincreased <- as.factor(baseball$payrankincreased)
```
## Machine Learning

### Hypothesis

$H_0$: $\beta_1$ = 0 - Teams payroll rank has no linear relationship with their winning %       
$H_0$: $\beta_1$ $\neq$ 0 - Teams payroll rank has a linear relationship with their winning % 

####Linear regression summary with confidence interval      
```{r primary, echo=FALSE}
# Here you can produce relevant tables and plots for your secondary objective
# The example below produces some plots using functions already defined in 
# another script
MLB <- lm(baseball$winpercent~baseball$payrank)
summary(MLB)

confint(MLB)

```
The P-value is practically 0 so there is a signifcant result and we can reject the null hypothesis and conclude that there is a linear realtionship between a teams winning % and where they rank in Team Payroll.

```{r, echo=FALSE}
winvsrank <- ggplot(baseball, aes(payrank, winpercent)) + geom_jitter() +  stat_smooth(method = "lm", col = "red") + labs(title = "Win % vs Payroll Rank", x = "Payroll Rank", y = "Win %") 
winvsrank + scale_x_continuous(breaks=c(5,10,15,20,25,30)) 
```

### Testing Linear Regression Assumptions

####Assessing Outliers        
```{r}
# Assessing Outliers

outlierTest(MLB) # Bonferonni p-value for most extreme obs
```
There are no outliers that would influence our model in a signifcant way.

####Normality of Residuals                               
```{r}
# Normality of Residuals

qqPlot(MLB, main="QQ Plot") #qq plot for studentized resid 
```

#### Distribution of studentized residuals

```{r}
# distribution of studentized residuals

sresid <- studres(MLB) 
hist(sresid, freq=FALSE, 
     main="Distribution of Studentized Residuals")
xfit<-seq(min(sresid),max(sresid),length=40) 
yfit<-dnorm(xfit) 
lines(xfit, yfit)

# plot studentized residuals vs. fitted values 
plot(fitted(MLB), residuals(MLB))
abline(h = 0)
plot(fitted(MLB), abs(residuals(MLB)))
summary(lm(abs(residuals(MLB))~fitted(MLB)))
```
After testing the residuals we can conclude that are data is an approximately normal distribution which is needed to fit are assumptions in our linear model. 

####Evaluation of homoscedasticity        
```{r}
# Evaluate homoscedasticity
# non-constant error variance test
ncvTest(MLB)
```
Since the P-value is .84 we can conclude that the standard deviations of the error terms are constant and do not depend on the x-value, which is an assumption needed for our linear model.

####Global validation of linear model assumptions .        
```{r}
gvmodel <- gvlma(MLB) 
summary(gvmodel)
```
All assumptions to our linear model are acceptable.

## Conclusion

The plot of team payroll rank and winning % showed a clear negative linear relationship. Meaning that as the Teams drop lower in the ranks of team payroll, their winning % is expected to also decrease. After performing all the diagnostic tests, the model passed each one confirming that it is a good model fit. The summary of the model showed a significant relationship between payroll rank and winning % at an $\alpha$ = .05 level, therefore, we can reject the null hypothesis $H_0$ and conclude, $H_a$ that there is a relationship between where a team ranks in team payroll and their winning %. According to our model for every spot a team drops in team payroll rank, their winning % is expected to decrease by 0.3%, and over a 162 game season that would be alomost 15 less wins expected  for the 30th ranked payroll vs the 1st ranked Payroll.