---
title: "Probability and Statistics"
author: "Will Pauls"
date: "February 19, 2019"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(Lahman)
library(dplyr)
library(tidyr)
library(readr)

# make team data frame with necessary variables
teams = Teams
teams = subset(teams, select = c(yearID, lgID, teamID, Rank, G, W, L, DivWin, WCWin)) 

# grab teams from 1985 and on to match with years of payroll information
teams = filter(teams, yearID >=1985)

# add winning % column
teams = mutate(teams, winpercent = round((W/(W+L)), digits = 3))

# add a playoffs column
teams$DivWin[is.na(teams$DivWin)] <- "N"
teams$WCWin[is.na(teams$WCWin)] <- "N"
teams <- mutate(teams, playoffs = if_else(DivWin == 'Y' | WCWin == 'Y', 'Y', 'N'))
teams = subset(teams, select = -c(DivWin, WCWin))
# add salaries to a data frame
sals = read_csv("Salaries.csv")         

# Get team payroll for each year
sals <- sals %>% 
  group_by(yearID, teamID) %>% 
    summarise(payroll = sum(salary))

#Standardize team names 
sals$teamID[sals$teamID=="CHN"] <- "CHC"
sals$teamID[sals$teamID=="CHA"] <- "CHW"
sals$teamID[sals$teamID=="KCA"] <- "KCR"
sals$teamID[sals$teamID=="LAN"] <- "LAD"
sals$teamID[sals$teamID=="NYN"] <- "NYM"
sals$teamID[sals$teamID=="NYA"] <- "NYY"
sals$teamID[sals$teamID=="SDN"] <- "SDP"
sals$teamID[sals$teamID=="SFN"] <- "SFG"
sals$teamID[sals$teamID=="SLN"] <- "STL"
sals$teamID[sals$teamID=="TBA"] <- "TBR"
sals$teamID[sals$teamID=="WSN"] <- "WAS"

teams$teamID <- as.character(teams$teamID)
teams$teamID[teams$teamID=="CHN"] <- "CHC"
teams$teamID[teams$teamID=="CHN"] <- "CHC"
teams$teamID[teams$teamID=="CHA"] <- "CHW"
teams$teamID[teams$teamID=="KCA"] <- "KCR"
teams$teamID[teams$teamID=="LAN"] <- "LAD"
teams$teamID[teams$teamID=="NYN"] <- "NYM"
teams$teamID[teams$teamID=="NYA"] <- "NYY"
teams$teamID[teams$teamID=="SDN"] <- "SDP"
teams$teamID[teams$teamID=="SFN"] <- "SFG"
teams$teamID[teams$teamID=="SLN"] <- "STL"
teams$teamID[teams$teamID=="TBA"] <- "TBR"
teams$teamID[teams$teamID=="WSN"] <- "WAS"
teams$teamID <- as.factor(teams$teamID)

# merge data into baseball
baseball <- merge(teams, sals, by=c("yearID", "teamID"))

baseball$teamID[baseball$teamID=="ML4"] <- "MIL"
baseball$yearID <- as.factor(baseball$yearID)
baseball$playoffs <- as.factor(baseball$playoffs)

# calculate Amount of payroll per team win
baseball <- baseball %>% mutate(dolperwin = payroll/W)

#calculate total payroll per year for entire MLB
mlb <- baseball %>% 
  group_by(yearID) %>% 
     summarise(mlbpayroll = sum(as.numeric(payroll)))

# get number of teams for each year
 numteams <-  baseball %>% 
  group_by(yearID) %>%
  summarise(numteams =length(yearID))

mlb$numteams <- numteams$numteams

# calculate average payroll per year
mlb <- mlb %>% 
  mutate(avgpayroll = mlbpayroll/numteams)
mlb <- subset(mlb, select = -c(numteams))

# merge into baseball data set
baseball <- merge(baseball, mlb, by="yearID")

# calculate percentange of total payroll each team has
baseball <- baseball %>% 
  mutate(percentofmlb = round((payroll/mlbpayroll), digits = 4))

#calculate average dollars of mlb payroll per win
wins <- baseball %>% 
  group_by(yearID) %>% 
  summarise(mlbwin= (sum(W)))

baseball <- merge(baseball, wins, by="yearID")
baseball <- baseball %>% 
  mutate(mlbdolperwin = mlbpayroll/mlbwin)

baseball <- baseball %>% group_by(yearID) %>% mutate(payrank = dense_rank((desc(payroll))))

# calculate teams difference in payroll from previous year
baseball<- baseball %>%
  arrange(teamID, yearID) %>%
  group_by(teamID) %>%
  mutate(paydiff = payroll - lag(payroll)) 

# +/- % change of payroll from previous year
baseball<- baseball %>%
  arrange(teamID, yearID) %>%
  group_by(teamID) %>%
  mutate(percentdiff = paydiff/lag(payroll)) 

baseball$percentdiff <- round(baseball$percentdiff, 4)

## calculate change in winning percentage from previous year
baseball<- baseball %>%
  arrange(teamID, yearID) %>%
  group_by(teamID) %>%
  mutate(winpercentdiff = winpercent - lag(winpercent)) 

## calculate change in percent of mlb from previous year
baseball<- baseball %>%
  arrange(teamID, yearID) %>%
  group_by(teamID) %>%
  mutate(percentofmlbdiff = percentofmlb - lag(percentofmlb)) 

## calculate change in payroll rank from previous year
baseball<- baseball %>%
  arrange(teamID, yearID) %>%
  group_by(teamID) %>%
  mutate(payrankdiff = payrank - lag(payrank)) 

# create column to test if payroll increased for team
baseball$payincreased[baseball$paydiff<=0]<- "No"
baseball$payincreased[baseball$paydiff>0]<- "Yes"

baseball$payincreased <- as.factor(baseball$payincreased)

# create a column to test if winning % increased for team
baseball$winincreased[baseball$winpercentdiff<=0]<- "No"
baseball$winincreased[baseball$winpercentdiff>0]<- "Yes"

baseball$winincreased <- as.factor(baseball$winincreased)

# create a column to test if percent of mlb increased for team
baseball$percentofmlbincreased[baseball$percentofmlbdiff<=0]<- "No"
baseball$percentofmlbincreased[baseball$percentofmlbdiff>0]<- "Yes"

baseball$percentofmlbincreased <- as.factor(baseball$percentofmlbincreased)

# create column to test if payroll increased for team
baseball$payrankincreased[baseball$payrankdiff<=0]<- "No"
baseball$payrankincreased[baseball$payrankdiff>0]<- "Yes"

baseball$payrankincreased <- as.factor(baseball$payrankincreased)
```
## Probability and Statistics

### Wins and Payroll Rank

We are looking to see if increasing where you rank in terms of team payroll is correlated with more winning. (This does not necessarily mean the amount a team paid in Payroll went up.)     

```{r}
# summary of teams whose winning increased after payroll rank increased
summary(baseball$winincreased[baseball$payrankincreased=="Yes"])

# summary of teams whose winning increased after payroll rank decreased
summary(baseball$winincreased[baseball$payrankincreased=="No"])
```
```{r, echo=FALSE}

# bar plot of win % increase and Payroll Rank Increase
barplot(matrix(c(189,183,248,264),nr=2), beside=T, 
        col=c("blue","red"), ylim = c(0,300), 
        names.arg=c("Payroll Rank Increased", "Payroll Rank Decreased"))
legend("topleft", c("Win % Increased","Win % Decreased"), pch=15, 
       col=c("blue","red"), 
       bty="n")

```

#### Hypothesis

$H_0$: $p_1$ = $p_2$ - The proportion of teams whose wins increased after increasing their Payroll Rank equals the proportion of teams whose winning increased after their payroll rank decreased.       
$H_a$: $p_1$ $\ge$ $p_2$ - The proportion of teams whose wins increased after increasing their Payroll Rank is greater than The proportion of teams whose winning increased after their payroll rank decreased. 
```{r}
payranktest <- prop.test(x = c(189, 248), n = c((189+183), (248+264)), alternative = "greater")
# Printing the results
payranktest 
```
Since the p-value is .2652 we do not have enough evidence to reject the null hypothesis that the two proportions are equal to each other.

### Wins and Payroll Amount

We are looking to see if increasing the amount of money you spend in Payroll in relation to other teams is correlated with winning


```{r}
# summary of teams whose winning increased after % of mlb pay increased
summary(baseball$winincreased[baseball$percentofmlbincreased=="Yes"])

# summary of teams whose winning increased after % of mlb pay decreased
summary(baseball$winincreased[baseball$percentofmlbincreased=="No"])
```
```{r, echo=FALSE}

# bar plot of win % increase and % of MLB Payroll Increase
barplot(matrix(c(224,248,213,199),nr=2), beside=T, 
        col=c("blue","red"), ylim = c(0,300),
        names.arg=c("% of MLB Payroll Increased", "% of MLB Payroll Decreased"))
legend("topleft", c("Win % Increased","Win % Decreased"), pch=15, 
       col=c("blue","red"), 
       bty="n")

```

#### Hypothesis

Since a lower proportion of teams have increased their winning after increasing their payroll than when they decrease in relation to other teams we will test both sides for this hypothesis.

$H_0$: $p_1$ = $p_2$ - The proportion of teams whose wins increased after increasing their Payroll amount in relation to other MLB teams equals the proportion of teams whose winning increased after their payroll amount decreased in relation to other teams.       
$H_a$: $p_1$ $\ne$ $p_2$ - The proportion of teams whose wins increased after increasing their Payroll amount in relation to other MLB teams does not equal the proportion of teams whose winning increased after their payroll amount decreased in relation to other teams.     

```{r}
payrolltest <- prop.test(x = c(224, 213), n = c((224+248), (213+199)))
# Printing the results
payrolltest
```
Since the p-value is .2337 we  do not have enough evidence to reject the null hypothesis that the two proportions are equal to each other.

### Conclusion

Looking at the proportions of teams whose winning increased after they increased their rank in payroll and after they increased their % of MLB payroll, it is apparent that the amount of money a team puts into their payroll isn't necessarily as important as where they rank in team payroll regardless of how much they spend. For this reason, our linear regrssion model we will use in the Machine Learning section will use Payroll Rank as the Independent variable. 